---
title: "Get started with tidyusmacro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with tidyusmacro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Every data function downloads full government flat files, so chunks are not
# evaluated at build time. This keeps builds fast and deterministic.
#
# `eval = FALSE` only governs rendering. R CMD check separately tangles this
# file with knitr::purl() and sources the result, and purl ignores `eval`
# (it respects `purl = FALSE`). Chunks that hit the network are therefore
# tagged `purl = FALSE` individually; a global setting here would not be seen,
# because purl does not evaluate this chunk. Chunks calling theme_esp() are
# tagged too: check sources the tangled file with echo = TRUE, so the plots
# print, and the "Public Sans" base family is not a font the pdf device can
# resolve. Everything else runs standalone, so the tangled file is a real
# smoke test.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(tidyusmacro)
```

tidyusmacro downloads full flat files from BLS, BEA, and FRED and returns them as tidy data frames, with the lookup tables already joined. One call gets you the entire release, which is what you want for exploratory work, in-depth research, and real-time analysis on data days. BLS and BEA files update right as releases go live; FRED usually follows about 40 minutes later.

## Data retrieval

### BLS flat files

`getBLSFiles()` downloads and processes data from Bureau of Labor Statistics flat files: CPI, CPI-W, chained CPI, PPI (commodity and industry), import/export prices, ECI, ECEC, productivity, JOLTS, CPS, CES, CEX, average prices, state/area employment (SAE), and local area unemployment (LAUS). Call `blsSources()` for the full, current list with sizes and descriptions. BLS asks for an email address in the user agent for flat-file downloads.

```{r, purl = FALSE}
cpi_data <- getBLSFiles(data_source = "cpi", email = "user@example.com")
jolts_data <- getBLSFiles(data_source = "jolts", email = "user@example.com")

# Newer sources use the same call; blsFiles() lists what's available within one
blsFiles("ppi", "user@example.com")
ppi_data <- getBLSFiles("ppi", "user@example.com")
```

Every source carries `freq` (`"monthly"`, `"quarterly"`, `"semiannual"`, or
`"annual"`) and `is_average` (`TRUE` for BLS-computed rows like the `M13`
annual average, which lands on December of that year, the same date as the
real December observation). These are not dropped by default -- pass
`include_averages = FALSE` to drop them, e.g. before grouping by date across
many series.

CPI arrives with monthly relative importance already attached, from BLS's
`cu.aspect` file: `weight`, the base for the observation month's 1-month
change, and `weight_12mo`, the base for its 12-month change -- the weights a
1-month and 12-month contribution calculation actually need, already shifted
to the right month (see `?getBLSFiles` for the dating convention; do not lag
either column yourself). Coverage is U.S. city average from March 2012
forward; weights are `NA` outside that rather than back-filled. They are keyed
on item code and month, not series ID, so seasonally adjusted series carry them
too. Pass `weights = FALSE` to skip the extra download.

```{r, purl = FALSE}
library(dplyr)

# Contribution of each item to the 12-month change in the all items index,
# in percentage points.
contributions <- cpi_data |>
  filter(area_code == "0000", seasonal == "U", periodicity_code == "R") |>
  arrange(series_id, date) |>
  group_by(series_id) |>
  mutate(contribution_12m = weight_12mo * (value / lag(value, 12) - 1)) |>
  ungroup()
```

`getCPIAspects()` returns the rest of the metadata in that file: BLS's own
contribution decomposition (`W1`, `WC`), median standard errors on the published
percent changes (`M1`, `MC`), and seasonal factors (`F`).

```{r, purl = FALSE}
effects <- getCPIAspects("user@example.com", aspect_type = c("W1", "WC"))
```

`getCESRevisions()` downloads the revisions table of the Current Employment Survey (CES) total jobs numbers straight from the BLS website.

```{r, purl = FALSE}
revisions_df <- getCESRevisions()
```

### BEA NIPA tables

`getNIPAFiles()` downloads and formats BEA NIPA data flat files, either monthly or quarterly values.

```{r, purl = FALSE}
nipa_quarterly <- getNIPAFiles(type = "Q")
nipa_monthly <- getNIPAFiles(type = "M")
```

`getPCEInflation()` loads and processes Personal Consumption Expenditures (PCE) inflation data with weights and growth measures.

```{r, purl = FALSE}
pce_monthly <- getPCEInflation("M")
pce_quarterly <- getPCEInflation("Q")
```

`getDallasTrimPCE()` builds the component-level panel underlying the Dallas Fed Trimmed Mean PCE inflation rate: monthly price changes, Fisher expenditure-share weights, and flags for which components are trimmed each month. Useful for replicating the trimmed-mean rate or analyzing what gets trimmed.

```{r, purl = FALSE}
# Default 24/31 Dallas Fed trim
panel <- getDallasTrimPCE()

# Replicate the monthly trimmed-mean rate
panel |>
  dplyr::filter(!is_trimmed) |>
  dplyr::group_by(date) |>
  dplyr::summarize(trim_pce = weighted.mean(price_change, weight))
```

### FRED

`getFRED()` downloads and merges economic data series from the Federal Reserve Economic Data (FRED) API.

```{r, purl = FALSE}
# Named arguments give friendly column names
fred_data <- getFRED(prime_epop = "LNS12300060", cpi = "CPIAUCSL")

# Unnamed arguments use lowercase ticker as column name
fred_data <- getFRED("UNRATE", "PAYEMS")
```

`getUnrateFRED()` is a convenience function to download unemployment level and labor force from FRED and calculate the unemployment rate.

```{r, purl = FALSE}
unrate_data <- getUnrateFRED()
```

## Statistical functions

`logLinearProjection()` performs log-linear projections on historical data. Designed for use within dplyr verbs.

```{r}
library(dplyr)

# Stands in for a real series; in practice this comes from getFRED() or
# getNIPAFiles(). Real GDP, billions of chained 2017 dollars, SAAR.
gdp_data <- data.frame(
  date = seq(as.Date("2015-01-01"), by = "quarter", length.out = 40),
  gdp  = 18000 * 1.005^(0:39)
)

# Fit the trend on 2015-2019 and project it through the end of the series.
gdp_data %>%
  mutate(projection = logLinearProjection(
    date = date,
    value = gdp,
    start_date = "2015-01-01",
    end_date = "2019-12-01"
  ))
```

## Visualization

The examples below use a small stand-in series rather than a download.

```{r}
library(ggplot2)

# `set.seed()` keeps the example reproducible.
set.seed(1)
plot_data <- data.frame(
  date  = seq(as.Date("2019-01-01"), by = "month", length.out = 60),
  value = 3.5 + cumsum(rnorm(60, sd = 0.1))
)

sector_data <- rbind(
  transform(plot_data, category = "Goods"),
  transform(plot_data, value = value + 1.2, category = "Services")
)
```

`theme_esp()` is a custom ggplot2 theme for Economic Security Project graphics with cream background and clean styling.

```{r, purl = FALSE}
ggplot(plot_data, aes(date, value)) +
  geom_line(color = esp_navy) +
  theme_esp()
```

`scale_color_esp()` and `scale_fill_esp()` provide ESP-branded color scales for ggplot2.

```{r, purl = FALSE}
ggplot(sector_data, aes(date, value, color = category)) +
  geom_line() +
  scale_color_esp() +
  theme_esp()
```

`date_breaks_gg()` creates intelligent date breaks for ggplot2 that always include the last data point.

```{r}
ggplot(plot_data, aes(date, value)) +
  geom_line() +
  scale_x_date(breaks = date_breaks_gg(n = 6, last = max(plot_data$date)))
```

`date_breaks_n()` generates evenly spaced date breaks by selecting every nth unique date.

```{r}
ggplot(plot_data, aes(date, value)) +
  geom_line() +
  scale_x_date(breaks = date_breaks_n(plot_data$date, n = 6))
```

## Included data

`cesDiffusionIndex` is a tibble with 250 rows mapping CES industry codes to industry titles.

```{r}
data(cesDiffusionIndex)
```

`dallasTrimPCEcomponents` is the 177-component dictionary used by `getDallasTrimPCE()`, mapping Dallas Fed trimmed-mean PCE components to BEA NIPA series codes and line numbers (Table 2.4.4U).

```{r}
data(dallasTrimPCEcomponents)
```
